Individual and Differential Harm in Redistricting

### 2022-05-04 paper/paper.Rmd
s5 finally

-   why counterfactuals?

    -   many factors influence the performance and qualities of an electoral system

    -   understanding how different components of the system contribute requires reaching in and manipulating those components

    -   just like causal inference, you are inherently interested in the counterfactual where some part of the electoral system is different

-   why representational harm?

    -   lit cites on why districts are good

    -   district-based electoral system working well if people are represented by the candidate of their choice

    -   this motivates *representational harm*

    -   we argue that harm is actually a great unifying framework for studying district-based electoral systems

@@

```{r mini-schematic}
#| fig.cap: |
#|      Schematic maps of the three running examples, 
#|      showing the partisan composition of each of the 36 precincts.
#|      The total voting population for each example is 180. 
#|      White lines delineate precincts.
include_graphics("figures/minis_schematic.pdf")
```

@@

```{r harm-mininois}
#| fig.cap: |
#|      Two potential districting plans for Mininois: 
#|      a Democratic gerrymander (a) and a close-to-proportional plan (b).
#|      Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual.
include_graphics("figures/harm_mininois.pdf")
```

Suppose now that plan (a) is enacted into law.
Who does this decision harm, if anyone?
We offer that a voter is harmed if they are not represented by their chosen party, but would have been so counterfactually.
Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
Republicans in the city and in the southwest are unharmed, as are all Democratic voters,, since they are represented by the same party under both plans.
Figure \ref{fig:harm-mininois}, Panel C illustrates this by plotting the number of voters harmed in each precinct.

@@

```{r diffharm-minissouri}
#| fig.cap: |
#|      Three potential districting plans for Minissouris: 
#|      a Democratic gerrymander (a), 
#|      a close-to-proportional plan (b), and a Republican gerrymander (c).
#|      Panels (d) and (f) show the total harm to voters in each precinct done 
#|      by their corresponding gerrymander, using (b) as a counterfactual.
#|      Panel (e) summarizes this harm by party and across precincts, showing 
#|      the average harm done to voters of each party by each gerrymander.
include_graphics("figures/diffharm_minissouri.pdf")
```

@@

[3] It's just different, but we also think it's better Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
Yet, as a collective, the approaches towards partisan symmetry suffer from a common issue: the lack of a counterfactual.

[4] specific issues with the key approaches and how we improve that.

[5] bring the mini-examples full circle with a little table of the scores and the harms?

## The Efficiency Gap

The efficiency gap ($EG$) interrogates how a party's votes are wasted under a districting plan.
By subtracting the wasted votes for Democrats from those for Republicans, and normalizing by the total votes cast, it offers a signed measure that is simple to interpret: $EG=0.05$, for example, indicates that 5% more Democratic votes are wasted than Republican ones.
The general recommendation is that an efficiency gap larger than two Congressional seats (i.e., 2 divided by the number of districts), or 0.08 for state legislative plans, indicates a partisan gerrymander.

A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
These concerns to not apply to our proposed measures.

First, $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$ are all averaged over possible election outcomes, and so do not suffer from discreteness.
And as discussed above, $U^{(G)}$ and $U^{(L)}$ both reward proportionality.
Of course, proportionality is not always possible, given a state's political geography.
The harm measures $H$ and $\Delta H(q)$ both reflect this by comparing to a set of counterfactual districting plans.
This addresses one inherent shortcoming of the efficiency gap, illustrated best by the state of Massachusetts.
According to the efficiency gap, Massachusetts' congressional districting plan is a Democratic gerrymander ($EG=-0.14$).
But it is impossible to draw a single Republican congressional district in the state.
Consequently $H$ and $\Delta H(q)$ are everywhere 0.

Finally, the fact that the efficiency gap is minimized for districts with 75%--25% vote shares is a direct consequence of comparing the raw number of wasted votes, rather than the share within each party.
In contrast, $\Delta H(q)$ is calculated by comparing harm averaged first within each party, avoiding this deficiency.

As we will see in the application below, often there is substantial correlation between the efficiency gap and $\Delta H(q)$.
Yet in more unusual or tricky cases, the efficiency gap may point to a partisan gerrymander while $\Delta H(q)$ indicates a fair outcome or an unfair advantage for the opposite party.

## Partisan Bias

Partisan bias is probably the most widely-used measure of partisan symmetry.
The bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
Like our proposed measures, this quantity is defined independent of electoral data and can be estimated using a uniform swing model.
Each value of $s$ yields information about a different point on the seats-votes curve.

In competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.

Fundamentally, however, $\beta(s)$ measures partisan symmetry, and not partisan harm or utility.
It should not be surprising, then, that it can and does "disagree" about the partisan effects of a particular plan.
These "disagreements" are really just differences in the target quantity of interest.

```{=html}
<!--
1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
2.  Focuses on only one part of S-V curve, which may not be relevant

Notably, as the quantity of interest is not partisan symmetry, our measures do not require the existence of a seats-votes curve. This is consequential for the study of congressional redistricting. As a seats-votes curve is defined for a legislature, rather than a delegation, to understand the impact of single state's congressional map, you would be forced to consider how every other state drew their maps [@katz:king:rose:21a]. Thus, if this taken as the definition of a seats-votes curve, then measures of partisan symmetry lack diagnostic power to identify gerrymanders for the delegations that academics, legal scholars, and the general public are often focused on. The individualized focus of our measures are, however, portable to congressional, state legislative, or any other form of districting and elections.
-->
```
## Mean-Median Difference

The mean-median difference straightforwardly measures the difference between the mean and median district vote shares.
In practice, the mean-median differences is moderately correlated with other partisan redistricting measures.
However, considerations of the mean and median district are only related to individuals insofar as they describe the general propensity to be harmed or benefited by a particular map.
As [@katz2020] discuss, the mean-median difference can be interpreted as an estimator of *votes-denominated bias*, another value derived from the seats-votes curve.

## Simulation Methods

Despite being derived from individual harm, differential partisan harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness.
In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry.

@@

```{r nj-maps, fig.width=6.5}
#| fig.cap: |
#|      Baseline vote patterns (a), two hypothetical gerrymanders (b) and (c), 
#|      and one randomly sampled plan (d).
#|      Census-designated places are shaded in gray and county lines in white.
include_graphics("figures/nj_maps.pdf")
```

@@

We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure \ref{fig:nj-maps}(d).
For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
[\^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.

@@

```{r nj-pairs, fig.width=7, fig.height=7}
#| fig.cap: |
#|      Scatterplots for all pairs of measures.
#|      The Democratic and Republican gerrymanders are shown 
#|      with the blue and red lines and crosses.
#|      The dashed line indicates a value of zero.
#|      Below the diagonal, numbers indicate the absolute value of the correlation between variables.
include_graphics("figures/nj_pairs.pdf")
```

@@

# Application to Voting Rights Litigation in Alabama {#sec:alabama}

```{r al-maps, fig.width=6.5}
#| fig.cap: |
#|      Baseline vote patterns (a), 
#|      fraction of the voting-age population which is Black (b),
#|      and fraction of voters in each precinct which are harmed by the 2020
#|      enacted congressional districts versus the set of four remedial plans (c).
#|      2020 enacted district boundaries are overlaid 
#|      in black on the left map and in white on the right maps.
include_graphics("figures/al_maps.pdf")
```

Differential racial harm: 0.23708

Differential partisan harm: 0.20939

Total harm: 0.089009

Average total harmed: 162390

Average total harmed by group:

| Average total harmed | Democratic | Republican |
|----------------------|------------|------------|
| **White**            | 28,476     | 7,867      |
| **Black**            | 125,635    | 147        |
| **Other**            | 257        | 7.4        |

: This table is just a placeholder to store the numbers.

[More text after the table.]

@@

```{r boston-map, fig.width=6.75, fig.height=3.375}
#| fig.cap: |
#|      Racial composition of the city of Boston (left), 
#|      and 2010 city coucil districts (right).
#|      Neighborhood boundaries are overlaid in black on the 
#|      left map and in white on the right map.
include_graphics("figures/boston_map.pdf")
```

The city of Boston is governed by a mayor and a nonpartisan 13-member city council.
Nine of the councillors are elected from contiguous city council districts, and the remaining four councillors are elected at-large, with all councillors serving two-year terms.
Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^4]
This section will study the extent to which the mixed at-large and district council system harms voters in comparison to a fully district-based system (as was the case before 1909 and from 1924-1951), either overall or differentially by racial group.
Figure \ref{fig:boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.

[^4]: City of Boston charter.
    Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.

@@

To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^5] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.

[^5]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.

@@

This paper is an attempt to bridge the quantitative gap between votes and representation.
We break from existing redistricting literature and focus on the relationship between votes and representation by copartisans, rather than the relationship between votes and seats.
This focus naturally leads to a new normative standard of partisan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters of each party.
This may occur regardless of whether the districting plan achieves partisan symmetry.

>>>>

@@

```{r mini-schematic}
#| fig.cap: |
#|      Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts.
#|      The total voting population for each example is 180. 
#|      White lines delineate precincts.
include_graphics("figures/minis_schematic.pdf")
```

@@

```{r harm-mininois}
#| fig.cap: |
#|      Two potential districting plans for Mininois: a Democratic gerrymander (a) and a close-to-proportional plan (b).
#|      Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual.
include_graphics("figures/harm_mininois.pdf")
```

Suppose now that plan (a) is enacted into law.
Who does this decision harm, if anyone?
We offer that a voter is harmed if they are not represented by their party of choice, but have a counterfactual where they would be represented by their party of choice.
Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
Republicans in the city and in the southwest are unharmed, as are all Democratic voters,, since they are represented by the same party under both plans.
Figure \ref{fig:harm-mininois}, Panel C illustrates this by plotting the number of voters harmed in each precinct.

@@

```{r diffharm-minissouri}
#| fig.cap: |
#|      Three potential districting plans for Minissouris: a Democratic gerrymander (a), 
#|      a close-to-proportional plan (b), and a Republican gerrymander (c).
#|      Panels (d) and (f) show the total harm to voters in each precinct done 
#|      by their corresponding gerrymander, using (b) as a counterfactual.
#|      Panel (e) summarizes this harm by party and across precincts, showing 
#|      the average harm done to voters of each party by each gerrymander.
include_graphics("figures/diffharm_minissouri.pdf")
```

@@

Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
The idea of partisan symmetry is rooted in assessing how parties would have performed under counterfactual voteshares.
Yet, this suffers from a core inflexibility: it does not allow for variation in districting.
Districting is the fundamental question of district-based democracies. 
*For whom can you vote? Who do you vote with?*
Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation. 
The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual.

Consider the major issues that have been raised over existing measures.
Partisan bias ($\beta(\half)$) can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
This has been called by some the Utah paradox [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
These stem from the omission of the alternatives.

By looking to alternatives, we ground the idea of zero in a balance of the harms, rather than a purely theoretical way.
This resolves the directionality problem common to other measures.
The use of a counterfactual in measuring the quantity of interest grounds the bias in the rules of what districts can be drawn.
This allows us to draw key strength from the impetus of the redistricting simulation literature.
Yet, the simplicity of the framework does not restrict only to the use of simulation methods.
Further, this framework does not need to rely on ideas of proportionality, which has no legal role in American democracy.
Nor does it suffer from discreteness in ensembles, as now both voteshares and districts lines shift.

Despite these differences, there are also key empirical similarities in many cases.
The efficiency gap and $\Delta H(q)$ are often quite well correlated under competitive states, as wasted votes and partisan harm are rooted in similar logics.
Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well. 
Where our framework rises above these though is especially in non-competitive states, where insurmountable advantages are rendered as zeroes and do not contribute to harm.

<!-- ## Simulation Methods -->

<!-- Despite being derived from individual harm, differential partisan harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below. -->
<!-- Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness. -->
<!-- In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes. -->
<!-- This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry. -->

@@

```{r nj-maps, fig.width=6.5, fig.height=3}
#| fig.cap: |
#|      Baseline vote patterns, two hypothetical gerrymanders, and one randomly sampled plan.
#|      Census-designated places are shaded in gray and county lines in white.
include_graphics("figures/nj_maps.pdf")
```

@@

We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure \ref{fig:nj-maps}(d).
For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
[^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.

@@

```{r nj-pairs, fig.width=7, fig.height=7}
#| fig.cap: |
#|      Scatterplots for all pairs of measures.
#|      The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.
#|      The dashed line indicates a value of zero.
#|      Below the diagonal, numbers indicate the absolute value of the correlation between variables.
include_graphics("figures/nj_pairs.pdf")
```

@@

@@

```{r boston-map, fig.width=6.75, fig.height=3.375}
#| fig.cap: |
#|      Racial composition of the city of Boston (left), and 2010 city coucil districts (right).
#|      Neighborhood boundaries are overlaid in black on the left map and in white on the right map.
include_graphics("figures/boston_map.pdf")
```

The city of Boston is governed by a mayor and a nonpartisan 13-member city council.
Nine of the councillors are elected from contiguous city council districts, and the remaining four councillors are elected at-large, with all councillors serving two-year terms.
Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^5]
This section will study the extent to which the mixed at-large and district council system harms voters in comparison to a fully district-based system (as was the case before 1909 and from 1924-1951), either overall or differentially by racial group.
Figure \ref{fig:boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.

[^5]: City of Boston charter.
    Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.

@@

To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^6] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.

[^6]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.

@@

For the menagerie, we proffer a new beast. *can we make the whole discussion a zoo metaphor*
This paper is an attempt to bridge the quantitative gap between votes and representation.
We break from existing redistricting literature and focus on the relationship between votes and representation by copartisans, rather than the relationship between votes and seats.
This focus naturally leads to a new normative standard of partisan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters of each party.
This may occur regardless of whether the districting plan achieves partisan symmetry.
